Unifying Gompertzian growth with the communicable disease spreading paradigm

### 2022-12-21 arxiv/main.tex
fix to emerges
update small fixes in main body

Zaporozhan and Ponomarenko (2010) proposed the cryptochrome protein-family to be the designated geomagnetic receptors able to pick up minute disturbances in magnetic fields which could translate into influenza symptoms \cite{zaporozhan2010mechanisms}. The cryptochrome proteins have also been linked to circadian regulation \cite{rodgers2009chemical} and have been extensively studied in the context of avian navigation \cite{hiscock2016quantum}. Besides the cryptochromes, Krylov (2017) presents evidence and references that another iron-containing protein called the magnetoreceptor protein (MagR) encoded by the gene CG8198 is of equal importance \cite{krylov2017biological}. He suggested that geomagnetic storms could be perceived by animals as a disruption of the precise synchronization of biological circadian rhythms with diurnal geomagnetic variation, an interplay involving the cryptochrome, MagR, and melatonin molecules.

Whatever the cause, global biosphere disturbance theory represents an alternative paradigm in disease spreading dynamics. Our goal with this study is to present a novel bridge between these two paradigms which at its core is a connection between the logistic growth model and the Gompertzian growth model. This connection represents an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible} and is instead a bridge built on higher order interaction terms between the susceptible and the infected populations. In doing this we shall show that the Gompertz model is incompatible with disease spreading through physical proximity interactions. Further, we show that the Gompertz curve emerge when individuals are connected through a common stressor affecting them simultaneously.

@@

\subsection*{The extended SIR model family is almost Gompertzian}
Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease SIR-like models. Under a double-log-transform of Eq. \ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve \cite{Levitt2020},
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow Carletti et al. (2020) \cite{carletti2020covid} and consider the extended version of the SIR model by including a group of deceased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This extension requires an addition to our original set of equations \ref{eq:SIR}, with an equation for the deceased group's growth at some rate $\delta$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

@@

In summary, we see that the SIR model family both lacks the observed immediate decrease in relative growth rate, and makes strong assumptions about the disease spreading network structure that imply non-local interactions. We are thus prompted to look for another model which can explain the observed mortality patterns.

@@

\section*{An alternative theory for observed mortality}
\label{seq:alt}
An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an instantaneous external stressor, initiating a stress response to eventually bring mortality rates back to stability.

@@

To build some intuition about these equations, we consider some simple scenarios. 
\begin{enumerate}
\item Let $x_j=1$, $\forall j \neq i$. In this scenario, Eq. \ref{eq:modSIR} becomes $\frac{d x_i}{dt} = - \alpha x_i$, implying as expected that the global biosphere disturbance is gone, and the value of $x_i$ goes to zero. In the communicable disease paradigm, Eq. \ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = \beta (1-x_i) - \alpha x_i$, and $x_i$ will grow as long as the right-hand side is positive. This is expected as all neighbors are able to transmit the disease.
\item Let $x_j \rightarrow 0$, $\forall j \neq i$. In this scenario, Eq. \ref{eq:modSIR} $\frac{1}{x_i}\frac{d x_i}{dt} \rightarrow \infty$, implying asymptotically infinite initial growth as the number of higher order terms increases. This behavior is in line with Gompertzian growth and is expected as growth rates are indicated by the susceptible. The implication is that the global biosphere disturbance has just started. In the communicable disease paradigm Eq. \ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = - \alpha x_i$, and the value of $x_i$ goes to zero due to the absence of infectious people.
\end{enumerate}
It is important to note that both sets of equations only describe the paradigm they represent. Consequently, the set of equations for one paradigm will appear nonsensical in the other paradigm. In the new paradigm, we do not consider an infected person as infectious, but rather consider the susceptible as the drivers of the growth, as they represent the stage of the global biosphere disturbance.

@@

\begin{equation}
\frac{d \ln{x}}{dt} = \sum_{i=1}^N s^i/i.
\end{equation}
Using standard integrals we see that the multiple terms amount to a Gompertz growth term adjusted by a Hypergeometric function,
\begin{eqnarray*}
  \sum_{i=1}^N s^i/i &=& \int_s \sum_{i=0}^{N-1} s^i ds\\
  &=&\int_s \frac{1-s^N}{1-s} ds\\
  &=&- \frac{{}_{1}F_{2}({1,N+1,N+2;s})}{N+1}s^{N+1} - \ln(1-s),
\end{eqnarray*}
where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. In summary, the modified logistic function emerges as
 \begin{equation}
 \label{eq:modLogistic}
\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} - \ln{x(t)},
 \end{equation}
 which represents the novel bridge between logistic and Gompertzian growth. Note that for $N=1$, this relative growth rate becomes the standard logistic function, while increasing the number of higher order terms will move the equation closer to the Gompertz curve as the Hypergeometric term approaches zero. This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig. \ref{fig:hypergeometric}). This novel growth equation can thus be included in the SIR equation for the infected to model collaborative growth.

\insertPdfFig{hypergeometric}{Comparison relative growth rate in the Richards $\theta$-logistic model (Eq. \ref{eq:rypdal}), Gompertz model (Eq. \ref{eq:medianGomp}), and the augmented causality-reversed SIR model (Eq. \ref{eq:modLogistic}), all as a function of the abundance variable. Both dependent and independent variables are normalized so that final abundance (or size) is set to unity. The Richards $\theta$-parameter is annotated in brackets in the legend, in addition to the number of terms used in the modified SIR model.}

@@

We can now further appreciate the Richards parametrization as a transition from non-collaborative to maximally collaborative growth as $\theta \rightarrow 0$. 
This feature of $\theta$ was also obtained by Petroni et al. (2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq. \ref{eq:rypdal} as non-linear resource availability dependent on the overall abundance, with linearity reached at $\theta=1$. The maximally coherent scenario implies that all units in the network are exposed to the stressor simultaneously.
Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \cite{morse1929diatomic}. 
In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields. The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.

@@

We further show that the observed mortality across countries can be explained by a model where the biosphere is stressed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established. The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between individuals or microscopic entities is present at all orders. From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems.

Thus, we see growth on a spectrum: In one extreme we find non-collaborative growth models or models with parameterized linear interaction effects, and in the other extreme we see a field of microscopic entities coherently sharing information much like quantum entangled particles. This observation adds to the list of emerging biological phenomena that exhibit coherent quantum behavior at the macroscopic level \cite{lambert2013quantum}. Although an global biosphere disturbance can explain our observations, one might still be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.

>>>>

Zaporozhan and Ponomarenko (2010) proposed the cryptochrome protein-family to be the designated geomagnetic receptors able to pick up minute disturbances in magnetic fields which could translate into influenza symptoms \cite{zaporozhan2010mechanisms}. The cryptochrome proteins have also been linked to circadian regulation \cite{rodgers2009chemical} and have been extensively studied in the context of avian navigation \cite{engels2014anthropogenic,
hiscock2016quantum}. Besides the cryptochromes, Krylov (2017) presents evidence and references that another iron-containing protein called the magnetoreceptor protein (MagR) encoded by the gene CG8198 is of equal importance \cite{krylov2017biological}. He suggested that geomagnetic storms could be perceived by animals as a disruption of the precise synchronization of biological circadian rhythms with diurnal geomagnetic variation, an interplay involving the cryptochrome, MagR, and melatonin molecules.

Whatever the cause, global biosphere disturbance theory represents an alternative paradigm in disease spreading dynamics. Our goal with this study is to present a novel bridge between these two paradigms which at its core is a connection between the logistic growth model and the Gompertzian growth model. This connection represents an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible} and is instead a bridge built on higher order interaction terms between the susceptible and the infected populations. In doing this we shall show that the Gompertz model is incompatible with disease spreading through physical proximity interactions. Further, we show that the Gompertz curve emerges when individuals are connected through a common stressor affecting them simultaneously.

@@

\subsection*{The extended SIR model family is almost Gompertzian}
Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease SIR-like models. Under a double-log-transform of Gompertzian growth Eq. \ref{eq:gomp_solution}, the growth curve is a simple straight line which is one of its characteristic traits \cite{Levitt2020},
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this discrepancy between the SIR-like models and the Gompertz model, we follow Carletti et al. (2020) \cite{carletti2020covid} and consider the extended version of the SIR model by including a group of deceased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This extension requires an addition to our original set of equations \ref{eq:SIR}, with an equation for the deceased group's growth at some rate $\delta$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

@@

In summary, we see that the SIR model family both lacks the observed immediate decrease in relative growth rate, and makes strong assumptions about the disease spreading network structure that imply non-local interactions. We are thus prompted to look for another model which can explain the observed mortality patterns. For this challenge, we shall first present a stochastic approach and then present a deterministic approach which unifies the SIR family models with the Gompetz curve.

@@

\section*{A stochastic theory for observed mortality}
\label{seq:alt}
An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an instantaneous external stressor, initiating a stress response to eventually bring mortality rates back to stability.

@@

To build some intuition about these equations, we consider some simple scenarios. 
\begin{enumerate}
\item Let $x_j=1$, $\forall j \neq i$. In this scenario, Eq. \ref{eq:modSIR} becomes $\frac{d x_i}{dt} = - \alpha x_i$, implying as expected that the global biosphere disturbance is gone, and the value of $x_i$ goes to zero. In the communicable disease paradigm, Eq. \ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = \beta (1-x_i) - \alpha x_i$, and $x_i$ will grow as long as the right-hand side is positive. This is expected as all neighbors are able to transmit the disease.
\item Let $x_j \rightarrow 0$, $\forall j \neq i$. In this scenario, Eq. \ref{eq:modSIR} becomes $\frac{1}{x_i}\frac{d x_i}{dt} \rightarrow \infty$, implying asymptotically infinite initial growth as the number of higher order terms increases. This behavior is in line with Gompertzian growth and is expected as growth rates are indicated by the susceptible. The implication is that the global biosphere disturbance has just started. In the communicable disease paradigm Eq. \ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = - \alpha x_i$, and the value of $x_i$ goes to zero due to the absence of infectious people.
\end{enumerate}
It is important to note that both sets of equations only describe the paradigm they represent. Consequently, the set of equations for one paradigm will appear nonsensical in the other paradigm. In the new paradigm, we do not consider an infected person as infectious, but rather consider the susceptible as the drivers of the growth, as they represent the stage of the global biosphere disturbance.

@@

\begin{equation}
\frac{d \ln{x}}{dt} = \sum_{i=1}^N s^i/i.
\end{equation}
Using standard integrals we see that the multiple terms amount to a Gompertz growth term adjusted by a Hypergeometric function,
\begin{eqnarray*}
  \sum_{i=1}^N s^i/i &=& \int_s \sum_{i=0}^{N-1} s^i ds\\
  &=&\int_s \frac{1-s^N}{1-s} ds\\
  &=&- \frac{{}_{1}F_{2}({1,N+1,N+2;s})}{N+1}s^{N+1} - \ln(1-s),
\end{eqnarray*}
where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. In summary, the modified logistic function emerges as
 \begin{equation}
 \label{eq:modLogistic}
\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} - \ln{x(t)},
 \end{equation}
 which represents the novel bridge between logistic and Gompertzian growth. Note that for $N=1$, this relative growth rate becomes the standard logistic function, while increasing the number of higher order terms will move the equation closer to the Gompertz curve as the Hypergeometric term approaches zero. This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig. \ref{fig:hypergeometric}). This novel growth equation can thus be included in the SIR equation for the infected.

\insertPdfFig{hypergeometric}{Comparison relative growth rate in the Richards $\theta$-logistic model (Eq. \ref{eq:rypdalGamma}), Gompertz model (Eq. \ref{eq:medianGomp}), and the augmented causality-reversed SIR model (Eq. \ref{eq:modLogistic}), all as a function of the abundance variable. Both dependent and independent variables are normalized so that final abundance (or size) is set to unity. The Richards $\theta$-parameter is annotated in brackets in the legend, in addition to the number of terms used in the modified SIR model.}

@@

We can now further appreciate the Richards parametrization as a transition from non-collaborative to maximally collaborative growth as $\theta \rightarrow 0$. 
This feature of $\theta$ was also obtained by Petroni et al. (2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq. \ref{eq:rypdalGamma} as non-linear resource availability dependent on the overall abundance, with linearity reached at $\theta=1$. The maximally coherent scenario implies that all units in the network are exposed to the stressor simultaneously.
Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \cite{morse1929diatomic}. 
In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields. The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.

@@

We further show that the observed mortality across countries can be explained by a model where the biosphere is disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established. The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between individuals or microscopic entities is present at all orders. From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems.

Thus, we see growth on a spectrum: In one extreme we find non-collaborative growth models or models with parameterized linear interaction effects, and in the other extreme we see a field of microscopic entities coherently sharing information much like quantum entangled particles. This observation adds to the list of emerging biological phenomena that exhibit coherent quantum behavior at the macroscopic level \cite{lambert2013quantum}. Although a global biosphere disturbance can explain our observations, one might still be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.

### 2022-12-21 letterToEditor/letter.tex
update letter
minor changes to sci per
update small fixes in main body

% personal data
\name{Matz Andreas Haugen}{}
\email{matzhaugen@gmail.com}
\address{Enerhauggata 1, 0651, Oslo, Norway}

@@

\recipient{To}{Editor, \\
               Chaos, Solitons, and Fractals,\\ 
               }
\date{\today}
\opening{\textbf{Initial submission of manuscript titled: ``Unifying the communicable disease spreading paradigm with Gompertzian growth''}}
\closing{Sincerely yours, \vspace{-1em}}

@@

If you do decide to review this paper you may find the enclosed list of possible referees helpful. Some of them are listed in the references of the manuscript. Due to the conclusions of our paper, I request that this paper receives more reviews than your average, and at least 3 reviews.

@@

\item Nicola Cufaro Petroni, cufaro@ba.infn.it,
Dipartimento di Matematica and TIRES, University of Bari (Ret).
Has done extensive work on the Gompertz model in relation to the Richard's model, connecting it with stochastic growth models and the Logistic (Lotka-Volterra) model. We use their concept of growth rooted in stochastic phenomena.
\item Salvatore De Martino,
demartino247@gmail.com, 
Dipartimento di Ingegneria dell’Informazione ed Elettrica.
Same as above
\item Marcin Molski, mamolski@amu.edu.pl, Adam Mickiewicz University, physics, Gompertz and quantum systems. Has done work on connecting the gompertz model to quantum coherent systems.
\item Francesco Piazza, Francesco.Piazza@cnrs-orleans.fr, Max Planck Institute for the Physics of Complex Systems, physics, SIR models.
Has written one of the central papers modeling covid disease propagation from the SIR perspective. We have based our methods on their paper.
\item Ernesto Estrada, estrada@ifisc.uib-csic.es, Institute of Interdisciplinary Physics and Complex Systems , mathematics, graph theory, SIR models.
Has written a recent paper on the Gompertz model in relation to communicable disease modelling including close connections with graph theoretical concepts used in our paper. 
\item K. Rypdal, kristoffer.rypdal@uit.no, University of Tromsø, mathematics, SIR models.
Has written a paper showing that SIR models become Gompertz in the extreme of the Richard's parameter. We also follow this paper's arguments in our paper.
\item Michael Levitt (NAS Member), michael.levitt@stanford.edu, Stanford University, computational biology, Gompertz models.
Has a preprint on the gompertz model in relation to covid where they leave the explanation of why mortality follows gompertz as an open question.
\item Chris G. Antonopoulos, canton@essex.ac.uk , U. of Essex, physics, chaos and SIR models.
Has written on the SIR model and chaotic dynamics.
\item Constantino Tsallis, tsallis@cbpf.br, Complexity Science Hub Vienna, Austria. Has done extensive work on long-range interaction and statistical mechanics. He might be able to see connections with his famous Tsallis entropy.

>>>>

% personal data
\name{Matz Andreas Haugen}{}
\email{matzhaugen@gmail.com}
\address{Valhallveien 13, 0196, Oslo, Norway}

@@

\recipient{To}{Editor, \\
               Bulletin of Mathematical Biology,\\ 
               }
\date{\today}
\opening{\textbf{Initial submission of manuscript titled: ``Unifying the communicable disease spreading paradigm with Gompertzian growth''}}
\closing{Sincerely yours, \vspace{-1em}}

@@

If you do decide to review this paper you may find the enclosed list of possible referees helpful. Some of them are listed in the references of the manuscript. Due to the conclusions of our paper, I request that this paper receives more reviews than your average, and at least 3 reviews. Some reviewers will probably have their scientific ``beliefs'' challenged. I would ask them to discern between dogma and observation. Science is only at the mercy of the latter.

@@

\item Vladimir N. Binhi, vnbin@mail.ru, Prokhorov General Physics Institute of the Russian Academy of Sciences, 38 Vavilov St., 119991 Moscow, Russia
Is familiar with the biological effects due geomagnetic disturbances and fields.
\item Andrei B. Rubin, rubin@biophysics.msu.ru, Faculty of Biology, Lomonosov Moscow State University, Leninskie Gory 1/12, 119234 Moscow, Russia
Is familiar with the biological effects due geomagnetic disturbances and fields.
\item Viacheslav V. Krylov, I.D. Papanin Institute for Biology of Inland Waters Russian Academy of Sciences, Borok, Russian Federation, krylovviacheslav@mail.ru
Is intimately familiar with geomagnetic disturbances' effect on biology.
\item Marcin Molski, mamolski@amu.edu.pl, Adam Mickiewicz University, physics, Gompertz and quantum systems. Has done work on connecting the gompertz model to quantum coherent systems.
\item Takuya Yamano, yamano@amy.hi-ho.ne.jp, Department of Physics, Ochanomizu University, 2-1-1 Otsuka, Bunkyo-ku, 112-8610, Tokyo, Japan
Is familiar with the gompertz model as it relates to statistical physics.
\item Shigui Ruan, ruan@math.miami.edu, Professor of Mathematics, University of Miami, 
An editor of Bulletin of Mathematical Biology
\item Francesco Piazza, Francesco.Piazza@cnrs-orleans.fr, Max Planck Institute for the Physics of Complex Systems, physics, SIR models.
Has written one of the central papers modeling covid disease propagation from the SIR perspective. We have based our methods on their paper.
\item Rober M May, robert.may@zoo.ox.ac.uk, Professor of Zoology, University of Oxford, 
\item Chris G. Antonopoulos, canton@essex.ac.uk, U. of Essex, physics, chaos and SIR models.
Has written on the SIR model and chaotic dynamics.
